Open source landscape ecology tools

### 2020-07-19 paper/paper.Rmd
adds some spatial data representations content

<!-- a paragraph or two about what are data representations, why do we need them, and what do they offer -->

<!-- two major data types - raster and vector -->

<!-- raster data representations -->
<!-- raster (+ mention terra) -->
<!-- stars -->

<!-- vector data representation -->
<!-- sp -->
<!-- sf -->
<!-- (conversion from text) -->
<!-- terra new representations -->

@@

<!-- access to spatial data -->

@@

<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->

>>>>

Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
Each geometry is connected to a row in an attribute table.
<!-- ... -->
A few hundred of different spatial data file formats were developed to store data in these formats.

While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????). <!--history section-->
Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
<!-- are pros/cons of that needed? -->
<!-- info about GDAL/PROJ? -->

The most prominent package allowing for raster data representation is *raster*.
<!-- maybe check number of reverse dependencies? -->
A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed, however, it could take several years for this package to be adopted by users.
Alternatively, *stars* package can be used to read and process raster data.
<!-- stars -->

The *sp* package was a standard for vector data representation for more than 10 years.
<!-- used by many packages -->
However, it is now in a maintenance mode only with a recommended successor - the *sf* package.
<!-- terra new vector representations -->
<!-- maybe check number of reverse dependencies? -->

Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
They also have dedicated methods or functions to convert between representations, for example between *sp* and *sf* objects or *raster* and *stars* objects.
This can be useful when a given function expects different data representation than currently existing.

@@

<!-- access to spatial data -->
<!-- rgee -->

@@

<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
<!-- rgee -->
